How to publish in Biological Sciences: a guide for the uninitiated

### 2023-03-31 docs/2.09-rebuttal.md

- Making a note of the line number where the revision is made (note that these can shift around in the revision)
- If you have reworded the text, do copy and paste that rewording into the rebuttal using quotes and corresponding line numbers
- Simply use a word like “done” to indicate changes on Minor comments
- Do be polite with your responses, but you don’t get any extra points for wordy thankfulness or praise. So keep it succinct and to the point
- Signal that you agree with the comment and that you have made a change to the text

Reviewers sometimes use a chatty style, and it may appear to you that they are asking you a question. For example, they might ask you exactly how accurate a piece of equipment you used to measure your organism. Intuitively, it seems like the right thing to do is to simply answer them in the rebuttal. But they expect you to make a change to the ms, and not simply to give them an answer in the rebuttal. Otherwise, it would have been pointless in making the comment.

Do bear in mind that your reviewer is a human, and was likely operating under less than ideal conditions when reading your manuscript. They could have been getting constant interruptions. They could have been reading it after having read another three manuscripts. They could suffer from insomnia and read it in the middle of the night with no sleep for a week. Give the reviewer the benefit of the doubt. Do remember to thank your reviewers and editors in your [acknowledgements](#acknowledgements). They’ve been working and doing the best for your manuscript without any thanks other than what you will give them. So give them a boost and help make their day that much brighter.

### Your rebuttal as an open document
Increasingly, journals are hosting the review process online with a DOI and are compelling contributors to have both reviews and rebuttals open for all to read. This is part of the opening of science and removal of bias and prejudices. It also means that  readers can pursue the sources of what they read and allows critical evaluation of the manuscript throughout the review process.

Ultimately, we might even ask whether we need an editorial decision of accept/reject if we are going to allow readers critical evaluation of an article. Surely, it is enough that the manuscript is reviewed? If authors and reviewers really are working together to produce the best manuscript, then as long as there are no fundamental flaws, the manuscript can be considered to have a default accepted status.

This brave new world is exactly where *eLife* has decided to go. Their plan is that once a paper has been accepted for review, there will be no ultimate editorial decision, only the movement between reviewers and authors to improve the manuscript between them. But this move did not happen without upset. Even though *eLife* had planned from the outset to go in this direction, some of the [gatekeepers](#boards) pushed back against the editor by calling for him to be fired. The people who really benefit from the current system which is heavy on prestige are the gatekeepers who have obtained and maintained their positions based on this prestige, something *eLife's* editor agrees on [@abbott2023strife].

This story reminds us that in order to change the publishing model we are going to be moving  against publishers who want to maintain their profits, and against [gatekeepers](#boards) who have gained so much from the current system. But this doesn't mean that it shouldn't be done. While *eLife* is a prominent and highly prestigious Biological Sciences publication, the vast majority of journals will also need to change, and this will require many smaller struggles repeated all over the world. With these changes, it will be imporatant to remember [why we publish](#whatjournalfor) and the [scientific project](https://howtowriteaphd.org/lifescientific.html) to which we aspire.

### 2023-04-05 1.00a-closedtoopen.Rmd
Add files via upload
# incluiding the CoI chapter

# The transition from closed to open {#closedtoopen}

There are a lot of problems in publishing in the biological sciences, but it doesn't have to be this way. The aim for this book is firstly to help you navigate the current 'closed' system (while acknowledging that parts are already open), and act as a catalyst toward a more open, transparent and equal system for the future not only of publishing science, but permeating throughout the culture of the scientific project. We have all the tools to make this transition now, and I think that this change will likely come within the time frame of the careers of you as an Early Career Researcher. But as you will see, this change needs to be driven.

While I'm going to pitch the transition from closed to open publishing as a simple process, and as a move from darkness into light, I acknowledge that it might best be defined as a [wicked problem](#wicked). I hope that a lot of these complexities will come out in the book, but acknowledge that a lot will be left unsaid. A large unanswered question is what happens to all the downstream impacts of losing a lucrative academic publishing business that employs tens of thousands of people. I like to think that many of these skilled people, who themselves are not recipients of the large fees acquired by the publishers, would be absorbed into the repurposed institutional libraries. No doubt, there will be casualties. But my belief is that the importance of the scientific project, and the [wicked problem](#wicked) we currently face in academic publishing, far outweighs the problems that we will see during the transition.

@@

Gatekeeping takes a lot of time and effort, and there are plenty of places where the current system lacks the transparency that is needed. In theory, there's nothing wrong with this model, were it not for the dark forces of for-profit publishers that have used the system for their own gain. Instead of the gatekeepers focussing on science, they have become distracted by metrics and hype promoted by publishers. Each of the three facets that surround the gatekeepers in the current publishing model (Figure \@ref(fig:publishing-triangle)) need to be changed to open up the system for a more equitable future, and eliminate the current biases that favour the select few.

### Is it possible to purge ourselves of the publishers? {#capture}

@@

There is a growing imperative to remove for-profit publishers from the academic system, and not just publishing. Just like social media companies, publishers have realised that data can be gleaned from academics and that this can be sold as a commodity [@posada2018inequality]. Moreover, academics produce data that can similarly be captured and sold. There is an entire work-flow from grant applications through to publishing that is up for grabs and the big publishing companies are already in the process of acquiring the various companies (typically offered as free tools to academics like Mendeley, Overleaf, Peerwith, Authorea, etc.) that they need to [capture, control and eventually dominate](#openness) and steer the entire process [@brembs2021replacing]: what Pooley calls 'surveillance publishing' [@pooley2021surveillance]. If we, as academics, want to remain independent of profit making companies (and in particular the big publishers), then we need to stop serving them now and move to [open platforms](#openscience).

@@

(ref:open-science) **A simple schematic for an Open Science publishing model**. Open science relies on the open nature of publication, communication and data. The gatekeepers still lie at the heart of this model, but are principally involved in ensuring the open and free flow of scientific information. This ideal world is free of the metrics that have dogged research in the past leading it into a profit making game for private publishers, and promoting the careers of a few elite scientists.

>>>>

# The transition from closed to open {#closedtoopen}
<br> 
There are a lot of problems in publishing in the biological sciences, but it doesn't have to be this way. The aim for this book is firstly to help you navigate the current 'closed' system (while acknowledging that parts are already open), and act as a catalyst toward a more open, transparent and equal system for the future not only of publishing science, but permeating throughout the culture of the scientific project. We have all the tools to make this transition now, and I think that this change will likely come within the time frame of the careers of you as an Early Career Researcher. But as you will see, this change needs to be driven.

While I'm going to pitch the transition from closed to open publishing as a simple process, and as a move from darkness into light, I acknowledge that it might best be defined as a [wicked problem](#wicked). I hope that a lot of these complexities will come out in the book, but acknowledge that a lot will be left unsaid. A large unanswered question is what happens to all the downstream impacts of changing a lucrative academic publishing business that employs tens of thousands of people. I like to think that many of these skilled people, who themselves are not recipients of the large fees acquired by the publishers, would be absorbed into the repurposed institutional libraries. No doubt, there will be casualties. But my belief is that the importance of the scientific project, and the [wicked problem](#wicked) we currently face in academic publishing, far outweighs the problems that we will see during the transition.

@@

Gatekeeping takes a lot of time and effort, and there are plenty of places where the current system lacks the transparency that is needed. In theory, there's nothing wrong with this model, were it not for some for-profit publishers that have used the system for their own gain. Instead of the gatekeepers focussing on science, they have become distracted by metrics and hype promoted by publishers. Each of the three facets that surround the gatekeepers in the current publishing model (Figure \@ref(fig:publishing-triangle)) need to be changed to open up the system for a more equitable future, and eliminate the current biases that favour the select few.

### Is it possible to do without the publishers? {#capture}

@@

There is a growing imperative to change the role of for-profit publishers in the academic system, and not just publishing. Just like social media companies, publishers have realised that data can be gleaned from academics and that this can be sold as a commodity [@posada2018inequality]. Moreover, academics produce data that can similarly be captured and sold. There is an entire work-flow from grant applications through to publishing that is up for grabs and the big publishing companies are already in the process of acquiring the various companies (typically offered as free tools to academics like Mendeley, Overleaf, Peerwith, Authorea, etc.) that they need to [capture, control and eventually dominate](#openness) and steer the entire process [@brembs2021replacing]: what Pooley calls 'surveillance publishing' [@pooley2021surveillance]. If we, as academics, want to remain independent of profit making companies (and in particular the big publishers), then we need to stop serving them now and move to [open platforms](#openscience) [@racimo2022Ethical].

#### Perspective on ethics of big publishers {#oil}
One of the very real problems that is associated with large publishing companies is that their business model is to make the maximum possible profits. This means that they will invest wherever they feel that they can make the most profit, and that includes promoting parts of their business that are involved with their investments. Elsevier is one of the big five publishers of scientific papers, but they are also heavily invested in oil and gas exploration and extraction [@westervelt2022Revealed; @macmillan2022Elsevier; @dahl2022AntiClimate]. This despite the fact that they are also publishing some of the best evidence that their own investments are causing irreparable damage to the planet: [Advances in Climate Change Research](https://www.sciencedirect.com/journal/advances-in-climate-change-research) [Energy and Climate Change](https://www.sciencedirect.com/journal/energy-and-climate-change), [Climate Change Ecology](https://www.sciencedirect.com/journal/climate-change-ecology), [The Lancet](https://www.thelancet.com) and many more titles that publish research on climate change. Moreover, Elsevier also publish 14 titles that aim to aid companies to extract more fossil fuels. It is not hard to see why more scientists are increasingly furious with Elsevier and have launched co-ordinated campaigns to boycott the company and highlight its hypocritical stance (visit: [Stop Elsevier](https://stopelsevier.wordpress.com/)). Elsevier can change, and if they start to feel the heat from enough scientists, then they will - just as they backed down over the pricing of their content with German universities [@schiermeier2018Germany]. These are not the only problems with [conflicts of interest](#conflicts) in scientific publishing (see [Part II](#oil2)) For other ways in which you can help stop the capture of science, see the [Last Note](#lastnote).

@@

(ref:open-science) **A simple schematic for an Open Science publishing model**. Open science relies on the open nature of publication, communication and data. The gatekeepers still lie at the heart of this model, but are principally involved in ensuring the open and free flow of scientific information. This ideal world is free of the metrics that have dogged research in the past.
